Kinematics of the Palomar 5 stellar stream from RR Lyrae stars

### 2019-08-22 paper/main.tex
Updates from Overleaf

\begin{abstract}
% Context
Thin stellar streams, formed from the tidal disruption of globular clusters, are important gravitational tools, sensitive to both global and small-scale properties of dark matter.
The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\sim 20\degr$ tidal tails connect back to the progenitor cluster, and the stream has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
However, most details of the phase-space distribution of the stream are not fully explained, and dynamical models that use the stream for other inferences are therefore incomplete.
% Aims
Here we aim to measure distance and kinematic properties along the Pal 5 stream in order to motivate improved models of the system. %Pal 5 stream.
We use a large catalog of RR Lyrae-type stars (RRLs) with astrometric data from the \Gaia\ mission to probabilistically identify RRLs in the Pal 5 stream.
RRLs are useful because they are intrinsically-luminous standard candles and their distances can be inferred with small relative precision ($\sim3\%$).
% Methods
% Results
By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find 25 RRL consistent with being members of the cluster (10) and stream (15).
Using these RRL, we detect gradients in distance and proper motion along the stream, and provide an updated measurement of the distance to the Pal 5 cluster using the RRL.
% Conclusions
We provide a catalog of Pal 5 RRLs with inferred membership probabilities for future modeling work.
\end{abstract}

@@

Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas:2001} in the QUEST survey; and more recently \citep{Ibata:2017} found about ten new RRLs using the catalogue of PS1 candidate RRLs from \citet{Hernitschek:2016}. At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia. Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalogue from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream. Both, key ingredients for dynamical modelling of the Pal~5 stream...
%meaning we can measure kinematic properties of the full stream...
\citep{Ibata:2017} found some RRL in pal 5, but no proper motions.
This is useful for modeling...
Many stream candidates found in RRL alone \citep{Mateu:2018}.

%---yes, I rewrite things many times. C.---
%RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence stars, but about $\sim 10,000$ time less numerous. At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia, so it's full extent can be traced and it's kinematic properties measured using RRLs. These stars can be reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.]{Holl2018,Drake2017,Mateu2012}, and are well known standard candles. Their distances can be estimated based on photometric data alone by means of their period-luminosity relation, with distance errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.  Previous studies have found RRLs in Pal5's tails: In the QUEST survey, \citet{Vivas2005} found 2 RRLs; while  \citep{Ibata:2017} found 12 RRLs using the catalogue of PS1 candidate RRLs from \citet{Hernitschek2015}. Combining \Gaia~DR2 proper motions with the newer PS1 RRL catalogue from \citet{Sesar2017b} --with improved classification probabilities and light curve periods--  will allow for precise distance and proper motion measurements along the full extent of the Pal~5 stream.
In this Article, we construct a probabilistic model for kinematic properties of RRLs in the vicinity of the Pal 5 stream to determine membership probabilities and thus map the distance and proper motion trends in the stream.
In \sectionname~\ref{sec:data}, we describe the source RR Lyrae catalog we use to search for Pal 5 members.
In \sectionname~\ref{sec:membership}, we implement a probabilistic membership model to identify candidate members of the stream using distance and astrometric information.
In \sectionname~\ref{sec:results}, we ...

@@

% From \Gaia~DR2 we have combined the two available RRL catalogues: the Specific Objects Study \citep[SOS][]{Clementini2018} and vari-classifier \citep[VC][]{Holl2018,Rimoldini2018}. The resulting VC+SOS catalogue contains a total of 228\,904, with all-sky coverage down to the mission's limiting magnitude $G=20.7$. After filtering out stars with contaminated photometry \verb+phot_bp_rp_excess_factor+>2 or null, as recommended in \citep{Rimoldini2018}, the VC+SOS catalogue contains a total of 175\,164 RR Lyrae. For the 122\,747 in SOS, which have undergone extensive light curve fitting, periods, amplitudes and other fitting parameters have been provided by \citet{Clementini2018}. The remaining 52\,417 stars (from the VC catalogue) did not have enough data points in \Gaia~DR2 to go into the SOS pipeline, so periods and other fitting parameters are not yet available for these stars.

To search for RRLs in the vicinity of Pal 5, we use a catalog of RRL stars from the PS1 survey that spans $\sim$3/4 of the sky (all sky north of declination $\delta > -30\degr$) and contains 61\,795 RRLs identified from their variability \citep{Sesar:2017b}.
In detail, the \emph{bona fide} RRLs in the catalog were classified using multi-epoch data in the $grizy$ photometric bands, with a typical number of 12 observations per filter, and has a limiting magnitude $r\sim21$ (corresponding to $G \sim 21$ for the typical colors of RRLs).
Periods and amplitudes in all five bands are available for all stars in this catalog.
We have chosen to use this catalog instead of the recently released \Gaia~RRL catalogs \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018} as these are subject to higher incompleteness in this particular area of the sky due to the \Gaia\ scanning law. Nevertheless, the vast majority (61\,625) of PS1's \emph{bona fide} RRLs have astrometric information available in \Gaia~\DR{2}.

\citet{Sesar:2017b} provide distances to the RRLs based on an $i$-band Period-Luminosity-Metallicity (PLZ) relation.
This has two key advantages when compared to optical bands: The PLZ relation in the $i$-band only weakly depends on metallicity, and the impact of dust extinction is less severe at longer wavelengths.
However, distances reported for the \rrc~stars in Table~5 of \citet{Sesar:2017b} have a systematic offset with respect to the \typeab~stars.
This offset comes from having applied the same PLZ to \typeab~and \typec~stars.
This is corrected by using the `fundamentalized' period in the PLZ relation for the \rrc~stars, which is computed as $\log{P_F} = \log P + 0.126$ \citep[following][]{Braga2016}.
In what follows, we use the $i$-band PLZ distances for the RRLs.
\citet{Sesar:2017b} estimate their overall distance precision to be 3\%, of which $\sim2$\% and $\sim1$\% correspond to systematic and random uncertainties, respectively.

@@

\citet{Sesar:2017b} provide an estimate of their catalog's completeness as a function of distance, at high galactic latitude, based on simulations. They estimate the mean completeness to be 92\% for \rrab~and 79\% for \rrc, up to a distance of 40~kpc, well beyond the distance of Pal~5 and its tails. At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%. We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogues to assess the completeness of both in a probabilistic manner. Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogues, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\textwidth]{members.pdf}
\caption{Sky positions of RRL...todo}
\label{fig:members}
\end{center}
\end{figure*}

@@

To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.
Correcting the observed number of 24~RRLs by the median completeness of 92\%, we get a total expected number of 26~RRLs, for which find an absolute magnitude $M_V=-8.1\pm 1.1$ \todo{CM: update this number based on 26}.
We therefore find that the initial mass of Pal 5 \todo{...}.

@@

\acknowledgments
This work was performed in part during the Gaia19 workshop and the 2019 Santa Barbara Gaia Sprint (also supported by the Heising-Simons Foundation), both hosted by the Kavli Institute for Theoretical Physics at the University of California, Santa Barbara. The Flatiron Institute is supported by the Simons Foundation. CM acknowledges funding from the MIA program at Universidad de la Republica, Uruguay.

>>>>

\begin{abstract}
% Context
Thin stellar streams, formed from the tidal disruption of globular clusters, are important gravitational tools, sensitive to both global and small-scale properties of dark matter.
The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\sim 20\degr$ tidal tails connect back to the progenitor cluster, and the stream has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
%SP the above sentence is a bit long
However, most details of the phase-space distribution of the stream are not fully explained, and dynamical models that use the stream for other inferences are therefore incomplete.
% Aims
Here we aim to measure distance and kinematic properties along the Pal 5 stream in order to motivate improved models of the system. %Pal 5 stream.
We use a large catalog of RR Lyrae-type stars (RRLs) with astrometric data from the \Gaia\ mission to probabilistically identify RRLs in the Pal 5 stream.
%SP throughout the manuscript, check if you mean plural when using RRL vs RRLs.
RRLs are useful because they are intrinsically-luminous standard candles and their distances can be inferred with small relative precision ($\sim3\%$).
% Methods
% Results
By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find 25 RRLs consistent with being members of the cluster (10) and stream (15).
Using these RRLs, we detect gradients in distance and proper motion along the stream, and provide an updated measurement of the distance to the Pal 5 cluster using the RRLs.
% Conclusions
We provide a catalog of Pal 5 RRLs with inferred membership probabilities for future modeling work.
\end{abstract}

@@

Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas:2001} in the QUEST survey; and more recently \citep{Ibata:2017} found about ten new RRLs using the catalog of PS1 candidate RRLs from \citet{Hernitschek:2016}. At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia. Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream. Both, key ingredients for dynamical modelling of the Pal~5 stream...
%meaning we can measure kinematic properties of the full stream...
\citep{Ibata:2017} found some RRL in pal 5, but no proper motions.
This is useful for modeling...
Many stream candidates found in RRL alone \citep{Mateu:2018}.

%---yes, I rewrite things many times. C.---
%RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence stars, but about $\sim 10,000$ time less numerous. At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia, so it's full extent can be traced and it's kinematic properties measured using RRLs. These stars can be reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.]{Holl2018,Drake2017,Mateu2012}, and are well known standard candles. Their distances can be estimated based on photometric data alone by means of their period-luminosity relation, with distance errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.  Previous studies have found RRLs in Pal5's tails: In the QUEST survey, \citet{Vivas2005} found 2 RRLs; while  \citep{Ibata:2017} found 12 RRLs using the catalog of PS1 candidate RRLs from \citet{Hernitschek2015}. Combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar2017b} --with improved classification probabilities and light curve periods--  will allow for precise distance and proper motion measurements along the full extent of the Pal~5 stream.
In this Article, we construct a probabilistic model for kinematic properties of RRLs in the vicinity of the Pal 5 stream to determine membership probabilities and thus map the distance and proper motion trends in the stream.
In \sectionname~\ref{sec:data}, we describe the source RR Lyrae catalog we use to search for Pal 5 members.
In \sectionname~\ref{sec:membership}, we implement a probabilistic membership model to identify candidate members of the stream using distance and astrometric information.
In \sectionname~\ref{sec:results}, we ...

@@

\todo{This part was rewritten to reflect the catalog we ended up using. CM.}

The PanSTARRS-1 (PS1) catalog of RRLs \citep{Sesar:2017b} contains 229K stars spanning $\sim$3/4 of the sky (all sky north of declination $\delta > -30\degr$), with multi-epoch $grizy$ data up to a limiting magnitude $r\sim21$ (corresponding to $G \sim 21$ for the typical colors of RRLs). These stars were identified as RRLs using machine-learning methods based on a template fitting algorithm that can find pulsation periods overcoming the sparse sampling of the PS1 survey ($\sim12$ epochs per filter).

Two all-sky RR Lyrae catalogs have also been released as part of \Gaia~\DR{2} \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018}. At present, however, these catalogs are subject to more significant and spatially-varying incompleteness in this particular area of the sky due to the \Gaia\ scanning law \citep[see][]{Rimoldini2018}.

Out of the 229K RRLs in the PS1 catalog, $\sim60$K are also reported as RRLs by \Gaia~\DR{2}, the vast majority of which (49K) are part of the subset defined by \citet{Sesar:2017b} as \emph{bona fide}. This subset is comprised of 61K RRLs above classification score thresholds set by \citet{Sesar:2017b} to ensure its high purity ($>90$\%) and completeness ($>80$\%). An extra $\sim6$K stars with lower classification scores in PS1 were also found to be RRLs by \Gaia~\DR{2} and have reported pulsation periods in the SOS catalog. Most of these RRLs have matching classifications and consistent periods in the two surveys, which increases our confidence that these are true RRLs.

For our present work, therefore, we use the superset of 68\,085 stars consisting of the 61\,795 \emph{bona fide} PS1 RRLs, plus the 6\,459 non-\emph{bona fide} PS1 stars also reported in the SOS \Gaia~\DR{2} catalog. Out of these, 68\,085 RRLs (99.8\%) have astrometric information available in \Gaia~\DR{2} and all---by construction---have photometric data and pulsation periods from PS1.

%distances---
\citet{Sesar:2017b} provide distances to the RRLs based on an $i$-band Period-Luminosity-Metallicity (PLZ) relation.
This has two key advantages when compared to optical bands: The PLZ relation in the $i$-band only weakly depends on metallicity, and the impact of dust extinction is less severe at longer wavelengths.
However, distances reported for the \rrc~stars in Table~5 of \citet{Sesar:2017b} have a systematic offset with respect to the \typeab~stars.
This offset comes from having applied the same PLZ to \typeab~and \typec~stars.
This is corrected by using the `fundamentalized' period in the PLZ relation for the \rrc~stars, which is computed as $\log{P_F} = \log P + 0.126$ \citep[following][]{Braga2016}.
In what follows, we use the $i$-band PLZ distances for the RRLs.
\citet{Sesar:2017b} estimate their overall distance precision to be 3\%, of which $\sim2$\% and $\sim1$\% correspond to systematic and random uncertainties, respectively.

@@

\citet{Sesar:2017b} provide an estimate of their catalog's completeness as a function of distance, at high galactic latitude, based on simulations. They estimate the mean completeness to be 92\% for \rrab~and 79\% for \rrc, up to a distance of 40~kpc, well beyond the distance of Pal~5 and its tails. At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%. We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogs to assess the completeness of both in a probabilistic manner. Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogs, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\textwidth]{members.pdf}
\caption{Sky positions of RR Lyrae stars with Pal 5 membership probabilities $>0.01$, colored by membership probability (large markers). The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.}
\label{fig:members}
\end{center}
\end{figure*}

@@

To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.  The $\log{N_{RR}-M_V$ has quite a large scatter so, particularly for such low numbers of RRLs, it will only allow us to give rough limits on the cluster's absolute magnitude. Correcting the observed number of 26~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which we find an absolute magnitude $M_V=-8.1\pm 1.1$, which correspond to the mode and 16/84th percentiles of the posterior probability PDF.
We therefore find that the initial mass of Pal 5 \todo{...}.

@@

\acknowledgments
This work was performed in part during the Gaia19 workshop and the 2019 Santa Barbara Gaia Sprint (also supported by the Heising-Simons Foundation), both hosted by the Kavli Institute for Theoretical Physics at the University of California, Santa Barbara. The Flatiron Institute is supported by the Simons Foundation. CM acknowledges funding from the MIA program at Universidad de la Republica, Uruguay, and is grateful for the hospitality and support of the Flatiron Institute where part of this research was carried out.
